Re-envisioning a future in scholarly communication

### 2017-05-30 manuscript.Rmd
Update to submitted version

---
title: "Re-envisioning a future in scholarly communication"
author: "Chris H.J. Hartgerink"
date: '`r Sys.Date()`'
output:
  html_document:
    includes:
      in_header: hypothesis.html
  word_document: default
fig_caption: yes
bibliography: library.bib
---

@@

Scholarly communication is in need of disruption. Commodifying knowledge as is currently done with journals, is not sustainable any longer. An alternative is the commodification of how information is consumed. By focusing on the commodification of consumption instead of commodification of the resource, the problem of access to knowledge can be resolved in a sustainable manner. Additionally, commodification of consumption removes several perverse incentives from the scholarly system that now produces unreliable knowledge. The main tenet underlying the themes of Open Access, Open Data, Open Science, and replication initiatives in scholarly communication is sustainability through transparency of the scholarly process in all facets. The sustainability of any networked system is threatened by single points of failure (i.e., the entire system can be manipulated from one node in the network). The scholarly process is ridden with such single points of failures at all stages. Distributing the scholarly communications system would remove the problems of single points of failure. Distributing and decentralizing the scholarly communications system is achievable with newly developed peer-to-peer (p2p) Internet protocols. Alongside decentralization and distribution of the content, integrity of the scholarly record can also be reformed to transform sections of a paper into different, reusable nodes of knowledge. These nodes can be logged on a blockchain based ledger of which everyone can have a copy. In order to deposit nodes onto the ledger, the depositor needs to agree that the contents are licensed CC 0, in order to maximize legal certainty regarding reuse of the contents. This is key to create a sustainable eco-system where scholars and companies can cooperate instead of compete, as we currently do.

@@

The scholarly process is much more complicated than can be captured in a single static and retrospective publication. The previously mentioned interactive elements enrich a single publication, but multiple moments of communicating ideas and results maps the empirical cycle onto the communication process. For example, a quantitative scholar needs to develop methods to investigate a research question. Feedback from peers would be valuable at that point. With so called Registered Reports [@10.1016/j.cortex.2012.12.016] a two-staged submission process is introduced to move away from a single static publication. It allows selected peers (i.e., peer reviewers) to discuss the contents at an interim stage. Additionally, this ensures chronology of predictions such as pre-registrations aim to do [@10.1038/s41562-016-0021], but in a more natural way. Nonetheless, registered reports, pre-registration, and data sharing are for the proactive, hence, remain the exception rather than the norm [@10.1371/journal.pbio.1002456], while these are already implicitly part of how scholarly research is conducted. Additionally, researchers create materials, protocols, and other outputs during the research process, share them with colleagues/advisors, but this process is not (aptly) sharable in a static publication. If the scholarly research process has more moments of communication, these implicit parts are finally able to show themselves.

Additionally, the scholarly process has grown more complicated with the rise of research tools available and the speed with which alternative results can be inspected. Point-and-click software has increased the number of statistical analyses a researcher can do within minutes. This increased researcher degrees of freedom to present results that are more in line with previous results, leading to confirmation bias or motivated reasoning [@10.1037/0033-2909.108.3.480], hence, invalid results [@10.1177/0956797611417632]. Moreover, the development of technology has created new types of data output that greatly exceed human comprehension and require algorithms before being able to make sense of the data (e.g., fMRI data). How a researcher applies all these tools is essential, but the current scholarly communication system does not require communication of the details needed to understand what steps were taken. Lack of direct reproducibility of results [e.g., @10.1101/056473] therefore is no surprise when we do not know how data are processed and analyzed. As such, the scholarly process has become more of a black box than it was before because communication has not adapted proportionate to the increasing complexity of the message that needs to be communicated. Increasing complexity and lack of transparency makes it increasingly difficult to reproduce steps taken, simply due to an increasing garden of forking paths [@10.1511/2014.111.460].

@@

Due to the fact that these complexities currently cannot and therefore are not reported, breaking the reign of the paper paradigm is a necessity to better represent and understand how scholarly results come into existence. Static paper publication remains a valid communication medium and it is not made obsolete when scholarly communications become truly digitally born and continuous. PDF versions of outputs can still be produced if a paper format is suitable or desirable. However, it is easy to imagine videos being embedded for protocols to be easily clarified (having "Video X" alongside "Figure X" and "Table X", for example) by origin, to then be replaced by stills when a paper version is produced. Nonetheless, without shifting the main focus from paper-based to digital-based scholarly communication, we cannot break free from the legacies of the paper paradigm. Paper-based communication includes the legacy of printing research outputs by default, which by extension includes the idea of subscriptions, purchasing content, and restricting the flow of information. Copyright is the legacy of the paper-based age that seems to limit the digital potential of scholarly outputs. As such, the legacy of a paper-based paradigm with respect to scholarly communication is that of knowledge commodification.

@@

Moreover, commodifying knowledge is not the only way to make scholarly communication sustainable; an alternative is the commodification of how information is consumed. That is, commodifying the services that serve the users the content, instead of the content itself. For example, if all scholarly information is freely available, (not-)for-profit organizations can compete for users who want to sift through that information efficiently. If an algorithm is developed that greatly increases efficiency of finding relevant articles, revenue could be built with usage fees to that service instead of the content. Such a market could create competition in information consumption, which is timely given the information explosion humanity is going through at the moment. The potential market capitalization of free flow of information and data is tremendous [e.g., freely reusable public sector information/data in the U.K. is estimated to provide a market cap of £590 million - £16 billion; p. 96 of @dotecon]. A freely reusable information resource would provide the infrastructure that encourages innovation and competition as to how people can discover and consume that information.

>>>>

---
title: "Re-envisioning a future in scholarly communication"
author: "Chris H.J. Hartgerink"
date: '`r Sys.Date()`'
output:
  html_document:
    includes:
      in_header: hypothesis.html
  pdf_document: default
  word_document: default
fig_caption: yes
bibliography: library.bib
---

@@

Scholarly communication is in need of disruption. Commodifying knowledge as is currently done with journals, is not sustainable any longer. An alternative is the commodification of how information is consumed. By focusing on the commodification of consumption instead of commodification of the resource, the problem of access to knowledge can be resolved in a sustainable manner. Additionally, commodification of consumption removes several perverse incentives from the scholarly system that now produce unreliable knowledge. The main tenet underlying the themes of Open Access, Open Data, Open Science, and replication initiatives in scholarly communication is sustainability through transparency of the scholarly process in all facets. The sustainability of any networked system is threatened by single points of failure (i.e., the entire system can be manipulated from one node in the network). The scholarly process is ridden with such single points of failures at all stages. Distributing the scholarly communications system would remove the problems of single points of failure. Distributing and decentralizing the scholarly communications system is achievable with newly developed peer-to-peer (p2p) Internet protocols. Alongside decentralization and distribution of the content, integrity of the scholarly record can also be reformed to transform sections of a paper into different, reusable nodes of knowledge. These nodes can be logged on a blockchain based ledger of which everyone can have a copy. In order to deposit nodes onto the ledger, the depositor needs to agree that the contents are licensed CC 0, in order to maximize legal certainty regarding reuse of the contents. This is key to create a sustainable eco-system where scholars and companies can cooperate instead of compete, as we currently do.

@@

The scholarly process is much more complicated than can be captured in a single static and retrospective publication. The previously mentioned interactive elements enrich a single publication, but multiple moments of communicating ideas and results maps the empirical cycle onto the communication process. For example, a quantitative scholar needs to develop methods to investigate a research question. Feedback from peers would be valuable at that point. With so called Registered Reports [@10.1016/j.cortex.2012.12.016] a two-staged submission process is introduced to move away from a single static publication. It allows selected peers (i.e., peer reviewers) to discuss the contents at an interim stage. Additionally, this ensures chronology of predictions such as pre-registrations aim to do [@10.1038/s41562-016-0021], but in a more natural way. Nonetheless, registered reports, pre-registration, and data sharing are for the proactive, hence, remain the exception rather than the norm [@10.1371/journal.pbio.1002456], while these are already implicitly part of how scholarly research is conducted. Additionally, researchers create materials, protocols, and other outputs during the research process, share them with colleagues/advisors, but this process is not (aptly) sharable in a static publication. If the scholarly research process has more moments of communication incorporating the many steps in research, these implicit parts are finally able to show themselves and allow credit to all types of researchers (e.g., theory builders, data-collectors, reanalyzers, etc.).

Additionally, the scholarly process has grown more complicated with the rise of research tools available and the speed with which alternative results can be inspected. Point-and-click software has increased the number of statistical analyses a researcher can do within minutes. This increased researcher degrees of freedom to present results that are more in line with previous results, leading to confirmation bias or motivated reasoning [@10.1037/0033-2909.108.3.480], hence, invalid results [@10.1177/0956797611417632]. Moreover, the development of technology has created new types of data output that greatly exceed human comprehension and require algorithms before being able to make sense of the data (e.g., fMRI data). How a researcher applies all these tools is essential, but the current scholarly communication system does not require communication of the details needed to understand what steps were taken, while at the same time neglecting the potential for education in (best) practices. Lack of direct reproducibility of results [e.g., @10.1101/056473] therefore is no surprise when we do not know how data are processed and analyzed. As such, the scholarly process has become more of a black box than it was before because communication has not adapted proportionate to the increasing complexity of the message that needs to be communicated. Increasing complexity and lack of transparency makes it increasingly difficult to reproduce steps taken, simply due to an increasing garden of forking paths [@10.1511/2014.111.460].

@@

Due to the fact that these complexities currently cannot and therefore are not reported, breaking the reign of the paper paradigm is a necessity to better represent and understand how scholarly results come into existence. Static paper publication remains a valid communication medium and it is not made obsolete when scholarly communications become truly digitally born and continuous. PDF versions of outputs can still be produced if a paper format is suitable or desirable. However, it is easy to imagine videos being embedded for protocols to be easily clarified (having "Video X" alongside "Figure X" and "Table X", for example) by origin, to then be replaced by stills when a paper version is produced. Nonetheless, without shifting the main focus from paper-based to digital-based scholarly communication, we cannot break free from the legacies of the paper paradigm. Paper-based communication includes the legacy of printing research outputs by default, which by extension includes the idea of subscriptions, purchasing content, and restricting the flow of information. Using copyright to limit access to scholarly communication is the legacy of the paper-based age that limits the digital potential of scholarly outputs. As such, the legacy of a paper-based paradigm with respect to scholarly communication is that of knowledge commodification.

@@

Moreover, commodifying knowledge is not the only way to make scholarly communication sustainable; an alternative is the commodification of how information is consumed. That is, commodifying the services that serve the users the content, instead of the content itself. For example, if all scholarly information is freely available, (not-)for-profit organizations can compete for users who want to sift through that information efficiently. If an algorithm is developed that greatly increases efficiency of finding relevant articles, this could generate revenue under various business models (e.g., subscription- or advertising model). Depending on the business model, this might shift inequality in access (current situation) to inequality in discovery. Nonetheless, a competitive market could create innovation in ways to consume information and puts pressure on the price levels to access these new services. New ways to consume information is timely given the information explosion humanity is going through at the moment. The potential market capitalization of free flow of information and data is tremendous [e.g., freely reusable public sector information/data in the U.K. is estimated to provide a market cap of £590 million - £16 billion; p. 96 of @dotecon]. A freely reusable information resource would provide the infrastructure that encourages innovation and competition as to how people can discover and consume that information.
